\textbf{The Illusion Game: A Novel Experimental Paradigm Provides Evidence in Favour of a General Factor of Visual Illusion Sensitivity and Personality Correlates}

### 2022-09-22 submission/manuscript.Rmd
other minor

  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com).
abstract: |
  Visual illusions highlight how the brain uses contextual and prior information to inform our perception of reality. Unfortunately, illusion research has been hampered by the difficulty of adapting these stimuli to experimental settings. In this study, we used the novel parametric framework for visual illusions to generate 10 different classic illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) varying in strength, embedded in a perceptual discrimination task. We tested the objective effect of the illusions on errors and reaction times, and extracted participant-level performance scores (n=250). Our results provide evidence in favour of a general factor (labelled Factor *i*) underlying the sensitivity to different illusions. Moreover, we report a positive relationship between illusion sensitivity and personality traits such as Agreeableness, Honesty-Humility, and negative relationships with Psychoticism, Antagonism, Disinhibition, and Negative Affect.
keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "3611"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

The parametric illusion generation framework developed in @makowski2021 proposes to conceptualize illusions as composed of targets and distractors that can be manipulated independently and continuously. In the present study, we have shown that such gradual modulation of illusion strength is effectively possible across 10 different types of classic visual illusions. Specifically, increasing the illusion strength led to an increase in the likelihood of errors, as well as the average and spread of RTs (but only up to a point, after which participants become faster at responding with the wrong answer). This important methodological step opens the door for new illusions-based paradigms and tasks to study the effect of illusions under different conditions and to quantify illusion sensitivity using objective behavioral outcomes - such as accuracy or speed - instead of subjective meta-cognitive reports. This new and complementary approach will hopefully help address some of the longstanding literature gaps, as well as cement illusions as valuable stimuli for the study of cognition.

>>>>

  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com).
abstract: |
  Visual illusions highlight how the brain uses contextual and prior information to inform our perception of reality. Unfortunately, illusion research has been hampered by the difficulty of adapting these stimuli to experimental settings. In this study, we used the novel parametric framework for visual illusions to generate 10 different classic illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) varying in strength, embedded in a perceptual discrimination task. We tested the objective effect of the illusions on errors and reaction times, and extracted participant-level performance scores (n=250). Our results provide evidence in favour of a general factor (labelled Factor *i*) underlying the sensitivity to different illusions. Moreover, we report a positive relationship between illusion sensitivity and personality traits such as Agreeableness, Honesty-Humility, and negative relationships with Psychoticism, Antagonism, Disinhibition, and Negative Affect.
keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "3628"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

The parametric illusion generation framework developed in @makowski2021 proposes to conceptualize illusions as composed of targets and distractors that can be manipulated independently and continuously. In the present study, we have shown that such gradual modulation of illusion strength is effectively possible across 10 different types of classic visual illusions. Increasing the illusion strength led to an increase in error likelihood, as well as the average and spread of RTs (but only up to a point, after which participants become faster at responding with the wrong answer). Using mixed models, we were able to statistically quantify the effect of illusions for each illusion and each participant separately. This important methodological step opens the door for new illusions-based paradigms and tasks to study the effect of illusions under different conditions and to measure illusion sensitivity using objective behavioral outcomes - such as accuracy or speed - instead of subjective meta-cognitive reports. This new and complementary approach will hopefully help address some of the longstanding literature gaps, as well as cement illusions as valuable stimuli for the study of cognition.

### 2022-09-22 submission/manuscript.tex
other minor

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{visual illusions, illusion game, Pyllusion, personality, general factor\newline\indent Word count: 3611}
\usepackage{lineno}

@@

The parametric illusion generation framework developed in Makowski et al. (2021) proposes to conceptualize illusions as composed of targets and distractors that can be manipulated independently and continuously. In the present study, we have shown that such gradual modulation of illusion strength is effectively possible across 10 different types of classic visual illusions. Specifically, increasing the illusion strength led to an increase in the likelihood of errors, as well as the average and spread of RTs (but only up to a point, after which participants become faster at responding with the wrong answer). This important methodological step opens the door for new illusions-based paradigms and tasks to study the effect of illusions under different conditions and to quantify illusion sensitivity using objective behavioral outcomes - such as accuracy or speed - instead of subjective meta-cognitive reports. This new and complementary approach will hopefully help address some of the longstanding literature gaps, as well as cement illusions as valuable stimuli for the study of cognition.

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{visual illusions, illusion game, Pyllusion, personality, general factor\newline\indent Word count: 3628}
\usepackage{lineno}

@@

The parametric illusion generation framework developed in Makowski et al. (2021) proposes to conceptualize illusions as composed of targets and distractors that can be manipulated independently and continuously. In the present study, we have shown that such gradual modulation of illusion strength is effectively possible across 10 different types of classic visual illusions. Increasing the illusion strength led to an increase in error likelihood, as well as the average and spread of RTs (but only up to a point, after which participants become faster at responding with the wrong answer). Using mixed models, we were able to statistically quantify the effect of illusions for each illusion and each participant separately. This important methodological step opens the door for new illusions-based paradigms and tasks to study the effect of illusions under different conditions and to measure illusion sensitivity using objective behavioral outcomes - such as accuracy or speed - instead of subjective meta-cognitive reports. This new and complementary approach will hopefully help address some of the longstanding literature gaps, as well as cement illusions as valuable stimuli for the study of cognition.
